The \proglang{R} Package \pkg{forestinventory}: Design-Based Global and Small Area Estimations for Multiphase Forest Inventories

### 2017-08-19 JStatSoft_forestinventory_master-concordance.tex
update Andreas 19.08.2017

\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:%
1 121 1 1 4 1 1 1 4 3 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Introduction.Rnw:ofs 128:%
1 95 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 224:%
136 34 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Main structure and modules of the Package.Rnw:ofs 259:%
1 84 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 344:%
172 15 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Twophase_Estimators and their Application.Rnw:ofs 360:%
1 124 1 1 2 4 0 1 2 14 1 1 5 7 0 1 2 2 1 1 2 24 0 1 2 4 1 1 2 4 0 1 2 2 1 1 2 1 0 1 5 6 0 1 2 3 1 1 2 8 0 1 2 156 1 1 7 6 0 1 1 26 0 1 2 3 1 1 2 13 0 1 2 3 1 1 6 5 0 1 1 8 0 1 2 5 1 1 6 5 0 1 1 8 0 1 2 7 1 1 2 4 0 1 2 1 1 1 6 5 0 1 1 3 0 1 2 1 1 1 2 11 0 1 2 3 1 1 6 5 0 1 1 8 0 1 2 4 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 888:%
189}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Threephase_Estimators and their Application.Rnw:ofs 889:%
1 92 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 982:%
191 10 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Calculation of Confidence Intervals.Rnw:ofs 993:%
1 13 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1007:%
203 8 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Special Cases and Scenarios.Rnw:ofs 1016:%
1 20 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1037:%
213 9 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Analysis and Visualization.Rnw:ofs 1047:%
1 20 1 1 4 6 0 1 2 2 1 1 6 8 0 1 2 2 1 1 7 9 0 1 2 2 1 1 6 8 0 1 2 2 1 1 7 9 0 1 2 4 1 1 5 7 0 1 2 6 1 1 2 4 0 1 2 2 1 1 2 24 0 1 2 16 1 1 4 1 1 1 2 11 0 1 2 79 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1289:%
224 20 1}

>>>>

\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:%
1 121 1 1 4 1 1 1 4 3 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Introduction.Rnw:ofs 128:%
1 95 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 224:%
136 34 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Main structure and modules of the Package.Rnw:ofs 259:%
1 78 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 338:%
172 15 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Twophase_Estimators and their Application.Rnw:ofs 354:%
1 124 1 1 2 4 0 1 2 14 1 1 5 7 0 1 2 2 1 1 2 24 0 1 2 4 1 1 2 4 0 1 2 2 1 1 2 1 0 1 5 6 0 1 2 3 1 1 2 8 0 1 2 156 1 1 7 6 0 1 1 26 0 1 2 3 1 1 2 13 0 1 2 3 1 1 6 5 0 1 1 8 0 1 2 5 1 1 6 5 0 1 1 8 0 1 2 7 1 1 2 4 0 1 2 1 1 1 6 5 0 1 1 3 0 1 2 1 1 1 2 11 0 1 2 3 1 1 6 5 0 1 1 8 0 1 2 4 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 882:%
189}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Threephase_Estimators and their Application.Rnw:ofs 883:%
1 213 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1097:%
191 10 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Calculation of Confidence Intervals.Rnw:ofs 1108:%
1 13 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1122:%
203 8 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Special Cases and Scenarios.Rnw:ofs 1131:%
1 20 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1152:%
213 15 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Analysis and Visualization.Rnw:ofs 1168:%
1 20 1 1 4 6 0 1 2 2 1 1 6 8 0 1 2 2 1 1 7 9 0 1 2 2 1 1 6 8 0 1 2 2 1 1 7 9 0 1 2 4 1 1 5 7 0 1 2 6 1 1 2 4 0 1 2 2 1 1 2 26 0 1 2 16 1 1 4 1 1 1 2 11 0 1 2 18 1 1 5 7 0 1 2 4 1 1 3 5 0 1 2 10 1 1 2 4 0 1 2 36 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1423:%
230 3 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Acknowledgements.Rnw:ofs 1427:%
1 4 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1432:%
235 11 1}

### 2017-08-19 rnw/Main structure and modules of the Package.Rnw
update Andreas 17.08.2017
update Andreas 19.08.2017

The two-phase or double-sampling estimators use inventory information from \textbf{two} nested samples which are commonly reffered to as \textit{phases} (figure \ref{fig:concmphase_and_sae_a}). The first phase $s_1$ comprises $n_1$ sample locations that each provide a set of explanatory variables described by the column vector $\pmb{Z}(x)\in{\Re^{p}}$ at each point $x \in s_1$. These explanatory variables are derived from auxiliary information that is available in high quantity within the forest area $F$. The second phase $s_2$ constitutes the terrestrial inventory conducted at $n_2$ subsamples of the large phase $s_1$ and provides the value of the target variable, i.e. the local density $Y(x)$ (e.g. the timber volume density per hectare). The set of explanatory variables at each sample location $x \in s_1$ is transformed into a prediction $\hat{Y}(x)$ of $Y(x)$ by the application of an oridinary least square (OLS) regression model. The basic idea of this setup is to boost the sample size by providing a large sample of less precise but cheap information about $Y(x)$ in $x \in s_1$, and to account for the lack of precision by collecting precise but costly information of $Y(x)$ at few sample locations $x \in s_2 \subset s_1$. In the \proglang{R} package \pkg{forestinventory}, estimators for two-phase sampling can be applied with the \code{twophase} function.

@@

The three-phase or triple-sampling estimators extend the principle of two-phase sampling and use inventory information from \textbf{three} nested samples (\textit{phases}) (figure \ref{fig:concmphase_and_sae_a}). The basic assumption is that auxiliary information is available in two different frequencies, i.e. sample sizes. The phase $s_0$ provides a large number $n_0$ of auxiliary information, whereas the phase $s_1$ provides additional auxiliary information that is only available at $n_1$ subsamples of $s_0$. The terrestrial information is then collected at a subsample $s_2$ of $s_1$ and $s_0$. The basic idea of three-phase sampling is that the additional set of auxiliary information $\pmb{Z}^{(2)}(x)\in{\Re^q}$ adds considerable explanatory power to the set of auxiliary information $\pmb{Z}^{(1)}(x)\in{\Re^p}$ available at all sample locations $x \in s_0$. From that it follows that we can define \textit{two nested} regression models. The \textit{full} set of predictor variables $\pmb{Z}^t(x)=(\pmb{Z}^{(1)t}(x),\pmb{Z}^{(2)t}(x))\in{\Re^{p+q}}$ in the sample $s_1$ can be used to calculate predictions $\hat{Y}_{1}(x)$ of $Y(x)$. The regression model applicable to the $s_1$ phase is thus referred to as the \textit{full model}. Whereas we can only use the subset of auxiliary information $\pmb{Z}^{(2)}(x)$ at the sample locations $s_0$, it allows for creating a much larger sample of less accurate predictions $\hat{Y}_{2}(x)$. If the gain in model precision from the reduced to the full model as well as the sampling fraction between $s_0$ and $s_1$ is sufficiently large, the three-phase estimator normally leads to a further increase in estimation precision compared to the two-phase estimator. In the \proglang{R} package \pkg{forestinventory}, estimators for three-phase sampling can be applied with the \code{threephase} function.

% NOTE:
% We still have to explain the difference between "non-exhausitve" and "exhaustive" auxiliary information (evtl. tricky)
% and the difference between cluster and no-cluster sampling
% --> I would not mentioned it here (keep  it general), but rather go into detail when describing the estimtors (next sections)

@@

We enter the topic of small area estimation if we want to make an estimation for a subregion $G$ of the entire inventory area $F$ (figure \ref{fig:concmphase_and_sae_b}). If the spatial extent of the subregion is small with respect to the terrestrial sample density, the number of terrestrial sample locations within the subregion $n_{2,G}$ will be to small to provide sufficient estimation accuracy. In this case, the application of multiphase estimation procedures can be efficient. However, the terrestrial sample size is too small to fit a separate regression model to them. The idea is then to \textit{borrow strength} from the entire terrestrial sample $s_2$ of $F$ and then to apply this  model to the small area. A potential model bias within $G$ is then corrected for by using the model residuals derived in the small area $G$. If no terrestrial samples fall into a small area $G$ (i.e. $n_{2,G}=0$), this bias correction cannot be applied and the resulting \textit{synthetic} estimates are potentially biased. Both two and three-phase small area estimators for synthetic and non-synthetic estimation can be applied by the \code{twophase} and \code{threephase} functions in the \proglang{R} package \pkg{forestinventory}.

@@

In the \proglang{R} package \pkg{forestinventory}, estimators for two-phase and three-phase sampling can be applied with the \code{twophase} and \code{theephase} function. From these two overall function calls, various estimators for specific inventory scenarios under the chosen sampling design can be applied (figure \ref{fig:struct_package}). The decision what estimator to use thereby follows a tree-like structure which can serve the user as a guideline. The first decision to make is whether an estimate and its variance should be computed for the entire inventory area (\textbf{global estimators}) or only for subregions of the entire inventory area (\textbf{small area estimators}). The next decision for both global and small area estimations is whether the explanatory variables were derived \textit{exhaustively} or \textit{non-exhaustively} over the inventory area. The first case allows for computing the exact mean $\bar{\pmb{Z}}$ of each explanatory variable over $s_1$ ($s_{1,G}$) and $s_0$ ($s_{0,G}$). In the second case, the mean of each explanatory variable over $s_1$ ($s_{1,G}$) and $s_0$ ($s_{0,G}$) can only be estimated by $\hat{\bar{\pmb{Z}}}$ which adds another component of uncertainty to the respective estimators. The package offers three small area estimators for both the exhaustive and non-exhaustive setup. The \textit{extended synthetic estimator} and the \textit{small area estimator} both apply a bias correction using the model residuals within small area $G$ and can only be applied if the terrestrial sample size $n_{2,G}$ within $G$ is $\geq 2$. The resulting point estimates and variances are asymptotically design-unbiased. If $n_{2,G}=0$, the bias correction can no longer be applied and the estimates of the repsective \textit{synthetic small area estimator} are potentially biased. The package also allows for calculating estimates solely based on the terrestrial sample $s_2$ (so called \textit{onephase estimation}) by the \code{onephase} function. All estimators are also available for \textit{cluster} sampling, in which case a sample unit conists of multiple, spatially agglomerated samples.

>>>>

The two-phase or double-sampling estimators use inventory information from \textbf{two} nested samples which are commonly reffered to as \textit{phases} (figure \ref{fig:concmphase_and_sae_a}). The first phase $s_1$ comprises $n_1$ sample locations that each provide a set of explanatory variables described by the column vector $\pmb{Z}(x)\in{\Re^{p}}$ at each point $x \in s_1$. These explanatory variables are derived from auxiliary information that is available in high quantity within the forest area $F$. The second phase $s_2$ constitutes the terrestrial inventory conducted at $n_2$ subsamples of the large phase $s_1$ and provides the value of the target variable, i.e. the local density $Y(x)$ (e.g. the timber volume density per hectare). The set of explanatory variables at each sample location $x \in s_1$ is transformed into a prediction $\hat{Y}(x)$ of $Y(x)$ by the application of an oridinary least square (OLS) regression model. The basic idea of this setup is to boost the sample size by providing a large sample of less precise but cheap information about $Y(x)$ in $x \in s_1$, and to account for the lack of precision by collecting precise but costly information of $Y(x)$ at few sample locations $x \in s_2 \subset s_1$. In the \proglang{R} package \pkg{forestinventory}, estimators for two-phase sampling can be applied with the \code{twophase()} function.

@@

The three-phase or triple-sampling estimators extend the principle of two-phase sampling and use inventory information from \textbf{three} nested samples (\textit{phases}) (figure \ref{fig:concmphase_and_sae_a}). The basic assumption is that auxiliary information is available in two different frequencies, i.e. sample sizes. The phase $s_0$ provides a large number $n_0$ of auxiliary information, whereas the phase $s_1$ provides additional auxiliary information that is only available at $n_1$ subsamples of $s_0$. The terrestrial information is then collected at a subsample $s_2$ of $s_1$ and $s_0$. The basic idea of three-phase sampling is that the additional set of auxiliary information $\pmb{Z}^{(1)}(x)\in{\Re^q}$ adds considerable explanatory power to the set of auxiliary information $\pmb{Z}^{(0)}(x)\in{\Re^p}$ available at all sample locations $x \in s_0$. From that it follows that we can define \textit{two nested} regression models. The \textit{full} set of predictor variables $\pmb{Z}^t(x)=(\pmb{Z}^{(0)t}(x),\pmb{Z}^{(1)t}(x))\in{\Re^{p+q}}$ in the sample $s_1$ can be used to calculate predictions $\hat{Y}_{1}(x)$ of $Y(x)$ at all sample locations $x \in s_1$. The regression model applicable to the $s_1$ phase is thus referred to as the \textit{full model}. Whereas we can only use the subset of auxiliary information $\pmb{Z}^{(0)}(x)$ in the so called \textit{reduced model} at the sample locations $x \in s_0$, it allows for creating a much larger sample of less accurate predictions $\hat{Y}_{0}(x)$. If the gain in model precision between the reduced and the full model as well as the sampling fraction between $s_0$ and $s_1$ is sufficiently large, the three-phase estimator normally leads to a further increase in estimation precision compared to the two-phase estimator. In the \proglang{R} package \pkg{forestinventory}, estimators for three-phase sampling can be applied with the \code{threephase()} function.

@@

We enter the topic of small area estimation if we want to make an estimation for a subregion $G$ of the entire inventory area $F$ (figure \ref{fig:concmphase_and_sae_b}). If the spatial extent of the subregion is small with respect to the terrestrial sample density, the number of terrestrial sample locations within the subregion $n_{2,G}$ will be to small to provide sufficient estimation accuracy. In this case, the application of multiphase estimation procedures can be efficient. However, the terrestrial sample size is too small to fit a separate regression model to them. The idea is then to \textit{borrow strength} from the entire terrestrial sample $s_2$ of $F$ and then to apply this  model to the small area. A potential model bias within $G$ is then corrected for by using the model residuals derived in the small area $G$. If no terrestrial samples fall into a small area $G$ (i.e. $n_{2,G}=0$), this bias correction cannot be applied and the resulting \textit{synthetic} estimates are potentially biased. Both two and three-phase small area estimators for synthetic and non-synthetic estimation can be applied by the \code{twophase()} and \code{threephase()} functions in the \proglang{R} package \pkg{forestinventory}.

@@

In the \proglang{R} package \pkg{forestinventory}, estimators for two-phase and three-phase sampling can be applied with the \code{twophase()} and \code{threephase()} function. From these two overall function calls, various estimators for specific inventory scenarios under the chosen sampling design can be applied (figure \ref{fig:struct_package}). The decision what estimator to use thereby follows a tree-like structure which can serve the user as a guideline. The first decision to make is whether an estimate and its variance should be computed for the entire inventory area (\textbf{global estimators}) or only for subregions of the entire inventory area (\textbf{small area estimators}). The next decision for both global and small area estimations is whether the explanatory variables were derived \textit{exhaustively} or \textit{non-exhaustively} over the inventory area. The first case allows for computing the exact mean $\bar{\pmb{Z}}$ of each explanatory variable over $s_1$ ($s_{1,G}$) and $s_0$ ($s_{0,G}$). In the second case, the mean of each explanatory variable over $s_1$ ($s_{1,G}$) and $s_0$ ($s_{0,G}$) can only be estimated by $\hat{\bar{\pmb{Z}}}$ which adds another component of uncertainty to the respective estimators. The package offers three small area estimators for both the exhaustive and non-exhaustive setup. The \textit{extended synthetic estimator} and the \textit{small area estimator} both apply a bias correction using the model residuals within small area $G$ and can only be applied if the terrestrial sample size $n_{2,G}$ within $G$ is $\geq 2$. The resulting point estimates and variances are asymptotically design-unbiased. If $n_{2,G}=0$, the bias correction can no longer be applied and the estimates of the repsective \textit{synthetic small area estimator} are potentially biased. The package also allows for calculating estimates solely based on the terrestrial sample $s_2$ (so called \textit{one-phase estimation}) by the \code{onephase()} function. All estimators are also available for \textit{cluster} sampling, in which case a sample unit consists of multiple, spatially agglomerated samples.

### 2017-08-19 rnw/Threephase_Estimators and their Application.Rnw
update Andreas 09.08.2017(4)
update Andreas 19.08.2017

% ---------------------------------------------------------------------------- %
\subsection{Global Estimators}

% - Merge Mathematical Background and Application
% - Quite Likely, we cannot go through all estimators ...
%   Idea: --> maybe only for non-exhaustive case, and then demonstrate with one example how the exhaustive can be applied in general
%
% - 
%
% - We could give the function call for all cases, but we should only show the output
%   for demonstrating particular things

% Concentrate on subset of estimators to go through the "tree": --> non-exhaustive case
% 1) nex, simple sampling, pseudo extended sae estimator
% 2) nex, simple sampling, pseudo sae estimator
% 3) nex, simple sampling, pseudo synthetic sae estimator --> give reason when to use (if n2G < 5, because ...)
% 
% 4) for the cluster case, show: nex, simple sampling, pseudo extended sae estimator
% 5) for the exhaustive case, show: ex, simple sampling, pseudo extended sae estimator
%    (mention that in cluster case, the cluster means shall be used)
% always use bw-adjustment (?)
%

\subsection{Small Area Estimators}

% The psynth estimator:
% 
% \begin{equation}\label{pseudosynth1}
% \hat{Y}_{G,psynth}=\hat{\bar{\pmb{Z}}}_{1,G}^t\hat{\pmb{\beta}}_{s_2}
% =\frac{1}{n_{1,G}}\sum_{x\in{s_{1,G}}}\hat{Y}(x)
% \end{equation}
% 
% \begin{equation}\label{estvarpseudosynth1}
% \hat{\var}(\hat{Y}_{G,psynth}): =
% \hat{\bar{\pmb{Z}}}_{1,G}^t\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}\hat{\bar{\pmb{Z}}}_{1,G}
% + \hat{\pmb{\beta}}_{s_2}^t\hat{\Sigma}_{\hat{\bar{\pmb{Z}}}_{1,G}}\hat{\pmb{\beta}}_{s_2}
% \end{equation}

>>>>

% ---------------------------------------------------------------------------- %
% ---------------------------------------------------------------------------- %
\subsection{Global Estimators}

% - Merge Mathematical Background and Application
% - Quite Likely, we cannot go through all estimators ...
%   Idea: --> maybe only for non-exhaustive case, and then demonstrate with one example how the exhaustive can be applied in general
%
% 
% NOTES:
% - mention that the exhaustive 3phase was originally called "2phase with partially exhaustive info"
% - clarify terminology "full" and "reduced"
%

% ------------------------------------- %
\subsubsection{Mathematical Background}

The the vector of regression coefficients for the \textbf{\textit{reduced model}} using the reduced set of explanatory variables $\pmb{Z}^{(0)}(x)$ is derived by solving the normal equation

\begin{eqnarray}\label{eq:normequ_redmod}
\hat{\pmb{\alpha}}_{s_2} &=& \Big(\frac{1}{n_2}\sum_{x\in{s}_2}\pmb{Z}^{(0)}(x)\pmb{Z}^{(0)t}(x)
\Big)^{-1}\frac{1}{n_2}\sum_{x\in{s}_2}Y(x)\pmb{Z}^{(0)}(x)
\end{eqnarray}

and likewise the vector of regression coefficients for the \textbf{\textit{full model}} using the full set of explanatory variables $\pmb{Z}^t(x)=(\pmb{Z}^{(0)t}(x),\pmb{Z}^{(1)t}(x))$:

\begin{eqnarray}\label{eq:normequ_redmod}
\hat{\pmb{\beta}}_{{s_2}}&=&\Big(\frac{1}{n_2}\sum_{x\in{s}_2}\pmb{Z}(x)\pmb{Z}^t(x)
\Big)^{-1}\frac{1}{n_2}\sum_{x\in{s}_2}Y(x)\pmb{Z}(x)
\end{eqnarray}

The variance-covariance matrices of the regression coefficients are thus defined as

\begin{subequations}\label{eq:covar3p}
\begin{align}
  \hat{\pmb{\Sigma}}_{\hat{\pmb{\alpha}}_{s_2}}=\Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{Z}^{(0)}(x)\pmb{Z}^{(0)t}(x) \Big)^{-1}
  \Big(\frac{1}{n_2^2}\sum_{x\in{s_2}}\hat{R}_{0}^2(x)\pmb{Z}^{(0)}(x)\pmb{Z}^{(0)t}(x)\Big)
  \Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{Z}^{(0)}(x)\pmb{Z}^{(0)t}(x) \Big)^{-1} \label{eq:covar_alpha} \\
  \hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}=\Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{Z}(x)\pmb{Z}^t(x) \Big)^{-1}
  \Big(\frac{1}{n_2^2}\sum_{x\in{s_2}}\hat{R}^2(x)\pmb{Z}(x)\pmb{Z}^t(x)\Big)
  \Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{Z}(x)\pmb{Z}^t(x) \Big)^{-1} \label{eq:covar_beta}
\end{align}
\end{subequations}

with the residuals $R_0(x)$ and $R(x)$ derived under the reduced and the full regression model respectively.

Likewise the two-phase estimation (section \ref{sec:twophase_and_appl}), also the three-phase estimation can be applied under \textit{non-exhaustive} and \textit{exhaustive} derivation of the explanatory variables at the $s_0$ phase.

In the \textbf{non-exhaustive case}, we use the \textit{estimated} means of $\pmb{Z}^{(0)}(x)$ over both the $s_0$ and $s_1$ phase, as well as the \textit{estimated} mean of $\pmb{Z}(x)$ over both the $s_1$ phase (equation \ref{meanvalues3pnex}). If the explanatory variables used in the reduced model are exhaustively derived (\textbf{exhaustive case}), the estimated mean $\hat{\bar{\pmb{Z}}}^{(0)}_0$ is simply replaced by the exact mean $\bar{\pmb{Z}}^{(0)}_0$. Note that in case of applied boundary adjustment (section \ref{sec:twophase_and_appl}), the simple mean is again replaced by the weighted mean analogous to equation \ref{eq:wmeanZ}.

\begin{equation}\label{meanvalues3pnex}
\hat{\bar{\pmb{Z}}}^{(0)}_0=\frac{1}{n_0}\sum_{x\in{s_0}} \pmb{Z}^{(0)}(x), \quad \hat{\bar{\pmb{Z}}}^{(0)}_1=\frac{1}{n_1}\sum_{x\in{s}_1}\pmb{Z}^{(0)}(x) ,
\quad \hat{\bar{\pmb{Z}}}_1=\frac{1}{n_1}\sum_{x\in{s}_1}\pmb{Z}(x)
\end{equation}

Under the internal model prerequisite which insures that the zero mean residual property over $F$, \textbf{point estimates} are then calculated as

\begin{subequations}\label{reg3p}
\begin{align}
\hat{Y}_{reg3p,ex}&=(\bar{\pmb{Z}}^{(0)}_0-\hat{\bar{\pmb{Z}}}^{(0)}_1)^t\hat{\pmb{\alpha}}_2 +
\hat{\bar{\pmb{Z}}}^t_1\hat{\pmb{\beta}}_{2} \label{eq:reg3p_ex} \\
\hat{Y}_{reg3p,nex}&=(\hat{\bar{\pmb{Z}}}^{(0)}_0-\hat{\bar{\pmb{Z}}}^{(0)}_1)^t\hat{\pmb{\alpha}}_2 +
\hat{\bar{\pmb{Z}}}^t_1\hat{\pmb{\beta}}_{2} \label{eq:reg3p_nex}
\end{align}
\end{subequations}

Equations \ref{eq:reg3p_nex} and \ref{eq:reg3p_ex} nicely demonstrate the philosophy of the three-phase estimation approach. Without the first term, the equations exactly equal the one of two-phase estimator where the estimated means in $\hat{\bar{\pmb{Z}}}_1$ are based on the $s_1$ sample and are transformed into predictions $\hat{Y}(x)$ of the local density $Y(x)$. The difference however is that we can now derive a more precise mean $\hat{\bar{\pmb{Z}}}^{(0)}_0$ (or even the exact mean $\bar{\pmb{Z}}^{(0)}_0$) over $s_0$ for a subset $\pmb{Z}^{(0)}(x) \in \pmb{Z}(x)$. The first term of equation \ref{eq:reg3p_nex} and \ref{eq:reg3p_ex} can thus be interpreted as a correction term that transforms the difference to the more precise / exact mean of $\pmb{Z}^{(0)}(x)$ into a residual of $Y(x)$ and adds it to the second term, thus making the point estimate more precise.

The \textbf{design-based variance} is then calculated as

\begin{subequations}\label{eq:dbvar_reg3p}
\begin{align}
\hat{\var}(\hat{Y}_{reg3p,ex})& =\frac{n_2}{n_1}\bar{\pmb{Z}}^{(0)t}\hat{\pmb{\Sigma}}_{\hat{\pmb{\alpha}}_{s_2}}
\bar{\pmb{Z}}^{(0)}+(1-\frac{n_2}{n_1})\hat{\bar{\pmb{Z}}}_1^t\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}
\hat{\bar{\pmb{Z}}}_1 \label{eq:dbvar_reg3p_ex} \\
\hat{\var}(\hat{Y}_{reg3p,nex})& =\hat{\pmb{\alpha}}_2^t\hat{\pmb{\Sigma}}_{\hat{\bar{\pmb{Z}}}^{(0)}_0}\hat{\pmb{\alpha}}_2
+\frac{n_2}{n_1}\hat{\bar{\pmb{Z}}}^{(0)t}_0
\hat{\pmb{\Sigma}}_{\hat{\pmb{\alpha}}_{2}}\hat{\bar{\pmb{Z}}}^{(0)}_0 + (1-\frac{n_2}{n_1})\hat{\bar{\pmb{Z}}}^{t}_1\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_2}\hat{\bar{\pmb{Z}}}_1 \label{eq:dbvar_reg3p_nex}
\end{align}
\end{subequations}

with the variance-covariance matrix $\hat{\Sigma}_{\hat{\bar{\pmb{Z}}}^{(1)}_0}$:

\begin{equation}\label{estvarcovaux3p}
\hat{\Sigma}_{\hat{\bar{\pmb{Z}}}^{(0)}_0}=
\frac{1}{n_{0}(n_{0}-1)}\sum_{x\in{s_{0}}}
(\pmb{Z}^{(0)}(x)-\hat{\bar{\pmb{Z}}}^{(0)}_{0})(\pmb{Z}(x)-\hat{\bar{\pmb{Z}}}^{(0)}_{0})^t
\end{equation}

We see that equation \ref{eq:dbvar_reg3p_ex} (exhaustive case) and \ref{eq:dbvar_reg3p_nex} (non-exhaustive case) differ by the term $\hat{\pmb{\alpha}}_2^t\hat{\pmb{\Sigma}}_{\hat{\bar{\pmb{Z}}}^{(0)}_0}\hat{\pmb{\alpha}}_2$ which is exactly the variance of the prediction $\hat{Y}_{0}(x)$ of the reduced model in the $s_0$ phase. This term can be made very small by chosing a large sample size $n_0$ for $s_0$ \citep{mandallaz2013techb}.

The package also gives the \textbf{external variance} defined as

\begin{subequations}\label{eq:extvar_reg3p}
\begin{align}
\hat{\var}(\hat{Y}_{reg3p,ex})&=\frac{1}{n_1}\frac{1}{n_2}\sum_{x\in{s_2}}\hat{R}_{0}^2(x))
+ (1-\frac{n_2}{n_1})\frac{1}{n^2_2}\sum_{x\in{s_2}}\hat{R}^2(x) \label{eq:extvar_reg3p_ex} \\
\hat{\var}_{ext}(\hat{Y}_{reg3p,nex})&=\frac{1}{n_0}\frac{\sum_{x\in{s_0}}(\hat{Y}_0(x)-\hat{\bar{Y}}_0)^2}{n_0-1} 
+\frac{1}{n_1}\frac{1}{n_2-1}\sum_{x\in{s}_2}(\hat{R}_0(x)-\hat{\bar{R_0}})^2 \nonumber\\
&+(1-\frac{n_2}{n_1})\frac{1}{n_2(n_2-1)}\sum_{x\in{s}_2}(\hat{R}(x)-\hat{\bar{R}})^2 \label{eq:extvar_reg3p_nex}
\end{align}
\end{subequations}

with ${R}(x)$ being the full model residuals calculated as $Y(x)-\pmb{Z}^t(x)\hat{\beta}_{s_2}$ and ${R}_{0}(x)$ being the reduced model residuals calculated as $Y(x)-\pmb{Z}^{(0)t}\hat{\alpha}_{s_2}$.

% \begin{subequations}\label{eq:var_reg3p_nex}
% \begin{align}
% \hat{\var}(\hat{Y}_{reg3p,nex})& =\hat{\pmb{\alpha}}_2^t\hat{\pmb{\Sigma}}_{\hat{\bar{\pmb{Z}}}^{(0)}_0}\hat{\pmb{\alpha}}_2
% +\frac{n_2}{n_1}\hat{\bar{\pmb{Z}}}^{(0)t}_0
% \hat{\pmb{\Sigma}}_{\hat{\pmb{\alpha}}_{2}}\hat{\bar{\pmb{Z}}}^{(0)}_0 + (1-\frac{n_2}{n_1})\hat{\bar{\pmb{Z}}}^{t}_1\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_2}\hat{\bar{\pmb{Z}}}^{t}_1 \label{eq:dbvar_reg3p_nex} \\
% \hat{\var}_{ext}(\hat{Y}_{reg3p,nex})& = \frac{1}{n_0}\frac{\sum_{x\in{s_0}}(\hat{Y}_1(x)-\hat{\bar{Y}}_1)^2}{n_0-1} \nonumber\\
% &+ \frac{1}{n_1}\frac{1}{n_2-1}\sum_{x\in{s}_2}(\hat{R}_1(x)-\hat{\bar{R_1}})^2+
% (1-\frac{n_2}{n_1})\frac{1}{n_2(n_2-1)}\sum_{x\in{s}_2}(\hat{R}(x)-\hat{\bar{R}})^2 \label{eq:extvar_reg3p_nex}
% \end{align}
% \end{subequations}

% ------------------------------------- %
\subsubsection{Application}

%--------------------------------------------------------------------------------------------------%
% ################################################################################################ %
%--------------------------------------------------------------------------------------------------%

% ---------------------------------------------------------------------------- %
\subsection{Small Area Estimators}

% ------------------------------------- %
\subsubsection{Mathematical Background}

%------------------------------------------------------------- %
\subsubsection{Application}

\newpage
